A New Method of Modelling Tuneable Lasers with Functional Composition

### 2020-11-15 Papers/Journal/Tuneable_Lasers.tex
Finished draft of nonlinear results

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{geometry}
\usepackage{xcolor,graphicx}
\usepackage[subfolder,cleanup]{gnuplottex}
%\usepackage{amsthm}
%\usepackage{enumitem}
%\usepackage{wrapfig}
\usepackage{subcaption}
%\usepackage{hyperref}
\usepackage{tikz}
\RequirePackage{cite}

@@

\section{Introduction}
\label{sec:intro}
Tuneable lasers have the ability to vary the frequency of its output by up to about 100 nanometres \cite{bohun2015, burgoyne2010, yamashita2009} and simultaneously lase at all frequencies within this bandwidth in short pulses. This tuneability leads to applications such as optical coherence tomography \cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy \cite{burgoyne2014}, deep tissue multi-photon microscopy \cite{chung2017, liu2017}, and diagnostics of ultrafast processes \cite{burgoyne2014, silfvast2004}. Tuneable lasers are usually constructed in a ring with five components: an output coupler, chirped fibre Bragg grating (CFBG), modulator, Er-doped fibre, and a pump laser. A typical tuneable ring laser cavity is depicted in Figure \ref{fig:cavity}. Due to the high power and short duratino of the pulses, the Kerr nonlinearity becomes crucial to the dynamics within the cavity.

The nonlinearity causes several effects to arise within the laser cavity due to the interplay of dispersion, modulation, and the nonlinearity \cite{bohun2015, coen1997, lapre2019, shao2019, woodward2018}. The effect of most interest for this paper is optical wave breaking which causes the leading edge of a pulse to be red shifted, and the trailing edge blue shifted\footnote{In the anomalous dispersion regime, the effect is reversed.} to the point that a shock is developed \cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The shift in frequencies cause the pulse to become rectangular in the frequency domain with a linear chirp over most of the pulse. Wave breaking manifests from self-phase modulation (SPM) which is a direct effect of the nonlinearity causing the pulse to interfere with itself. This interference generates new frequencies which induces the rectangular profile \cite{agrawal2013, mollenauer1980, woodward2018}. Modulation instablility is another effect that will be of interest that typically arises in the anomalous dispersion regime. However, in the presence of multiple pulses modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation (XPM) and four-wave mixing (FWM) \cite{agrawal1987, agrawal1989, agrawal2013, haelterman1992}. Modulation instability causes a wave to break into short pulses, and can cause a laser to lose coherence \cite{agrawal1987, coen1997, haelterman1992}. Within a ring laser cavity wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. This gives rise to the need of understanding the rich landscape of the parameter space, and determining design principles to guarantee the ring laser is stable and sustainable \cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

*** Paragraph explaining method / results / layout

\section{Modelling Efforts}
\label{sec:modelling}
The standard equation for studying nonlinear optics is the nonlinear Schr\"odinger equation (NLSE),
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
	\label{eq:smallnlse}
\end{align}
Here $A = A(T, z) : \mathbb{R}^2 \mapsto \mathbb{C}$ is the complex pulse amplitude, $\beta_2 \in \mathbb{R}$ is the second order dispersion, and $\gamma \in \mathbb{R}$ is the coefficient of nonlinearity. In practice, the NLSE, \eqref{eq:smallnlse}, lacks a few key terms, thus, it is often generalized by adding amplification and loss\footnote{Occasionally higher order terms are added as well.}. These additions give the generalized nonlinear Schr\"{o}dinger equation (GNLSE) \cite{agrawal2013, bohun2015, finot2008, peng2018, shtyrina2017, yarutkina2013},
	\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
	\label{eq:nlse}
\end{align}
where $g(A)$ is an amplifying term due to the gain, and $\alpha \in \mathbb{R}^+$ is the loss due to scattering and absorption.

@@

To proceed we must choose the form of the amplification term, $g(A)$. We assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},& E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, and $\Es$ is the energy at which the gain begins to saturate \cite{haus1984, shtyrina2017, silfvast2004}. This reduces the GNLSE, \eqref{eq:nlse}, to
\begin{align}
	\pdiff{A}{z} &= \frac{g_0 A}{2 \left( 1 + E / \Es \right)}.
\end{align}
By assuming $E(0) = E$ and $E(L_g) = E_{\text{out}}$, we find the energy after amplification is
\begin{align}
	\label{eq:engain}
	E_{\text{out}} &= \Es W \left( \frac{E}{\Es} \textrm{e}^{E / \Es} \textrm{e}^{g_9 L_g} \right),
\end{align}
where $W$ is the Lambert $W$ function \cite{corless1996}. Rewriting \eqref{eq:engain} in terms of the pulse gives
\begin{align}
	G(A) &= \left( \frac{\Es}{E} W \left( \frac{E}{\Es} \textrm{e}^{E/\Es} \textrm{e}^{g_0 L_g} \right) \right)^{1/2} A
\end{align}
for the amplification of the pulse within the Er-doped fibre. Repeating this process for the loss, dispersion, and nonlinearity yields
\begin{align}
	L(A) &= (1 - R) \textrm{e}^{- \alpha L_T}A, \\
	D(A) &= \FTi{\textrm{e}^{i \omega^2 L_D\beta_2/2} \FT{A}}, \\
	F(A) &= A \textrm{e}^{i \gamma |A|^2 L_f},
\end{align}
where $R$ is the reflectivity of the output coupler\footnote{Depending on the layout of the laser cavity the loss may instead take the form $L(A) = R \textrm{e}^{- \alpha L_T}A$.}, $L_T$ is the total length of the laser circuit, $L_D$ is the length of the dispersive medium, $L_f$ is the length of fibre between the amplifier and the output coupler, and where $\mathcal{F}$ denotes the Fourier transform. Finally, we must assume a form for the modulation. The modulation is considered to be applied externally in which ever way the operator sees fit. For simplicity the representation is taken as the Gaussian
\begin{align}
	M(A) &= \textrm{e}^{-T^2 / 2 T_M^2} A,
	\label{eq:modform}
\end{align}
where $T_M$ is the characteristic width of the modulation. This assumption is not particularly restrictive. Calcaterra and Boldt have shown that Gaussians can form a basis for $L^2(\mathbb{R})$ \cite{calcaterra2008a}. Therefore, we can approximate any modulation function in $L^2(\mathbb{R})$ by a sum of \eqref{eq:modform}.

@@

Now that we have the algebraic effect of each section of the cavity, we are ready to compose each map together to give the effect of one round trip of the cavity. Thus, we must now consider the order of the components. As we are most interested in the output of the laser cavity, we shall start with the loss. We then pass the pulse through the CFBG followed by the modulator. The pulse then travels through the Er-doped fibre to be amplified, and finally we consider the effect of the nonlinearity. The nonlinearity must immediately follow the gain---this is the section of the cavity where the pulse is most energetic, and hence, has the largest effect. Note, the permutation of the other components is indeed important as the operators do not, in general, commute---this is in contrast to the models described in Section \ref{sec:discrete}. Moreover, we prefer the loss to follow the gain and nonlinearity as we wish to minimize the length over which the nonlinearity has an effect, and maximize the output power. Functionally we denote one round trip of the cavity by
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A))))).
\end{align}
 The pulse after one complete circuit of the laser cavity is then returned back into the cavity to restart the process. A steady solution to this model is one in which the envelope and chirp are unchanged after traversing every component in the cavity---we are uninterested in the phase. That is, such that $\mathcal{L}(A) = A \textrm{e}^{i \phi}$---for some $\phi \in \mathbb{R}$. An example of the evolution of the envelope during one round trip of the laser cavity can be found in Figure \ref{fig:cavityevo}.

\section{Results}
We split the results into two subsections. In the following subsection we investigate the low nonlinearity limit, and in Section \ref{sec:nlresults} we look at the full nonlinear model.

@@

\begin{figure}[tbp]
	\includegraphics{Sample_Gauss}
	\caption{Solution to the linear model given by \eqref{eq:A0}.}
	\label{fig:samplegauss}
\end{figure}

@@

\subsection{Nonlinear Solution and Instability}
\label{sec:nlresults}

In this iterative scheme---as well as within the laboratory---we must specify the input pulse shape. The most common form is a hyperbolic secant \cite{coen1997, finot2008, mitschke1986, rothenberg1989b, tomlinson1984}, which we assume has the exact form
\begin{align}
	A_0 = \Gamma \sech \left( 2 T \right) \textrm{e}^{i \pi / 4},
	\label{eq:nlA0}
\end{align}
where $\Gamma$ is a normalizing factor chosen so the pulse has an initial energy of $E_0 = 0.1$, unless specified otherwise. We plot the evolution of a pulse through the cavity in Figure \ref{fig:evolution}. We find very different behaviour in these two examples despite very similar parameters. In Figure \ref{fig:convevo} the input pulse is very noisy, but, sheds off the left lobe and reaches an equilibrium state quickly after. In the opposite extreme, Figure \ref{fig:breakevo} has the smooth initial pulse given in \eqref{eq:nlA0}, however, the nonlinearity causes the pulse to pulsate or `breathe'. This effect compounds with each round trip, which ultimately results in the degradation of the pulse into an unusable and unrecoverable state. The only difference between these two realizations, besides the initial pulse shape, is the value of the nonlinearity parameter, $b$.

stability

instability

energy

error

permutation

chirp / chirp Saturation

shape / kurtosis

\begin{figure}[tb]
	\centering
	\begin{subfigure}{\columnwidth}
		\centering
		\includegraphics{Conv}
		\caption{Example of a pulse stabilizing from a seemingly random initial pulse shape ($b = 1.0$).}
		\label{fig:convevo}
	\end{subfigure} \\
	\begin{subfigure}{\columnwidth}
		\centering
		\includegraphics{Break}
		\caption{Example of a pulse destabilizing from a hyperbolic secant initial pulse ($b = 1.6$).}
		\label{fig:breakevo}
	\end{subfigure}
	\caption{Two extremes of the evolution of a pulse ($s = 0.1$, $E_0 = 0.1$, $a = 8000$, $h = 0.04$).}
	\label{fig:evolution}
\end{figure}

\newpage

In the nonlinear case, we recover this linear response for moderate values of $T$. However, for $|T| > 1$ the chirp begins to saturate. This is consistent with the experimental results \cite{chen, rothenberg, tomlinson}.

Wave breaking is not limited to just optics; wave breaking occurs in areas such as plasmas, transmission lines, and fluid dynamics \cite{rothenberg}. Wave breaking occurs because the pulse begins to interfere with itself in a way called SPM \cite{agrawal2002, agrawal2013, becker}. SPM occurs because the index of refraction is intensity dependent \cite{agrawal2002, becker, rothenberg, silfvast}, which leads to additional chirp across the pulse \cite{agrawal2013, anderson, rothenberg, silfvast}. This in turn causes higher order frequencies to be injected into the pulse \cite{agrawal2013, anderson},

We obtain comparable results as in the experiments \cite{anderson, rothenberg}

To obtain a better understanding of how the pulse either converges to equilibrium, or diverges to wave breaking, we shall examine the difference between the envelopes of consecutive iterations. More precisely, we compute the error by
\begin{align}
	\textrm{E} = \frac{\| |A_i| - |A_{i-1}| \|_2}{\| A_{i-1} \|_2},
	\label{eq:error}
\end{align}
where $\| \cdot \|_2$ denotes the $L^2(\mathbb{R})$ norm, which is computed numerically using the trapezoid rule ($N = 2^{18}$).

he standard method would be to iterate until a fixed tolerance is reached, however, there are some reasons that make a fixed number of iterations preferable---as long as a sufficient number is chosen. The main reason is that this allows us to observe a richer structure than simply whether or not the tolerance had been reached by some maximum number of iterations. Additionally, incorrectly choosing the critical tolerance could easily lead to erroneous categorizations at points.

The last item we wish to consider is the order in which the components are placed.

xpm \cite{agrawal2013, agrawal1987, agrawal1989}

\begin{figure}[tbp]
	\centering
	\includegraphics{ChirpFT}
	\caption{}
	\label{fig:chirpft}
\end{figure}

\begin{figure}[tbp]
	\centering
	\includegraphics{ParamSpaceErr}
	\caption{}
	\label{fig:error}
\end{figure}

\begin{figure}[tbp]
	\centering
	\includegraphics{ParamSpaceEnergy}
	\caption{}
	\label{fig:}
\end{figure}

\begin{figure}[tbp]
	\centering
	\includegraphics{ParamSpaceEnergySwitch}
	\caption{}
	\label{fig:energyswitch}
\end{figure}

\section{Conclusion}

>>>>

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{geometry}
\usepackage{xcolor,graphicx}
%\usepackage[subfolder,cleanup]{gnuplottex}
%\usepackage{amsthm}
%\usepackage{enumitem}
%\usepackage{wrapfig}
\usepackage{subcaption}
%\usepackage{hyperref}
\usepackage{tikz}
\usepackage{cite}

@@

\section{Introduction}
\label{sec:intro}
Tuneable lasers have the ability to vary the frequency of its output by up to about 100 nanometres \cite{bohun2015, burgoyne2010, yamashita2009} and simultaneously lase at all frequencies within this bandwidth in short pulses. This tuneability leads to applications such as optical coherence tomography \cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy \cite{burgoyne2014}, deep tissue multi-photon microscopy \cite{chung2017, liu2017}, and diagnostics of ultrafast processes \cite{burgoyne2014, silfvast2004}. Tuneable lasers are usually constructed in a ring with five components: an output coupler, chirped fibre Bragg grating (CFBG), modulator, Er-doped fibre, and a pump laser. A typical tuneable ring laser cavity is depicted in Figure \ref{fig:cavity}. Due to the high power and short duration of the pulses, the Kerr nonlinearity becomes crucial to the dynamics within the cavity.

The nonlinearity causes several effects to arise within the laser cavity due to the interplay of dispersion, modulation, and the nonlinearity \cite{bohun2015, coen1997, lapre2019, shao2019, woodward2018}. The two effects of most interest for this paper are optical wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to be red shifted, and the trailing edge blue shifted (in the anomalous dispersion regime, the effect is reversed) to the point that a shock is developed \cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The shift in frequencies cause the pulse to become rectangular in the frequency domain with a linear chirp over most of the pulse. In optics, wave breaking manifests from self-phase modulation (SPM) which is a direct effect of the nonlinearity causing the pulse to interfere with itself. This interference generates new frequencies which induces the rectangular profile \cite{agrawal2013, mollenauer1980, woodward2018}. Outside of optics, wave breaking occurs in other areas with nonlinear waves such as plasmas, heat propagation, transmission lines, and fluid dynamics \cite{rothenberg1989b}. Modulation instability is another effect that will be of interest that typically arises in the anomalous dispersion regime. However, in the presence of multiple pulses modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation (XPM) and four-wave mixing (FWM) \cite{agrawal1987, agrawal1989, agrawal2013, haelterman1992}. Modulation instability causes a wave to break into short pulses, and can cause a laser to lose coherence \cite{agrawal1987, coen1997, haelterman1992}. Within a ring laser cavity wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. This gives rise to the need of understanding the rich landscape of the parameter space, and determining design principles to guarantee the ring laser is stable and sustainable \cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

In Section \ref{sec:modelling}, we briefly review previous mathematical models for the evolution of laser pulses, then we derive our own model in Section \ref{sec:model}. We then examine the results of our model in Section \ref{sec:results}---first solving the model analytically assuming the absence of the nonlinearity, then numerically solving the full nonlinear model. Finally, the concluding thoughts and possible directions of future work are given in Section \ref{sec:conclusion}.

\section{Modelling Efforts}
\label{sec:modelling}
The standard equation for studying nonlinear optics is the nonlinear Schr\"odinger equation (NLSE),
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
	\label{eq:smallnlse}
\end{align}
Here $A = A(T, z) : \mathbb{R}^2 \mapsto \mathbb{C}$ is the complex pulse amplitude, $\beta_2 \in \mathbb{R}$ is the second order dispersion, and $\gamma \in \mathbb{R}$ is the coefficient of nonlinearity. In practice, the NLSE, \eqref{eq:smallnlse}, lacks a few key terms, thus, it is often generalized by adding amplification and loss (occasionally higher order terms are added as well). These additions give the generalized nonlinear Schr\"{o}dinger equation (GNLSE) \cite{agrawal2013, bohun2015, finot2008, peng2018, shtyrina2017, yarutkina2013},
	\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
	\label{eq:nlse}
\end{align}
where $g(A)$ is an amplifying term due to the gain, and $\alpha \in \mathbb{R}^+$ is the loss due to scattering and absorption.

@@

To proceed we must choose the form of the amplification term, $g(A)$. We assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},& E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, and $\Es$ is the energy at which the gain begins to saturate \cite{haus1984, shtyrina2017, silfvast2004}. This reduces the GNLSE, \eqref{eq:nlse}, to
\begin{align}
	\pdiff{A}{z} &= \frac{g_0 A}{2 \left( 1 + E / \Es \right)}.
\end{align}
By assuming $E(0) = E$ and $E(L_g) = E_{\text{out}}$, we find the energy after amplification is
\begin{align}
	\label{eq:engain}
	E_{\text{out}} &= \Es W \left( \frac{E}{\Es} \textrm{e}^{E / \Es} \textrm{e}^{g_9 L_g} \right),
\end{align}
where $W$ is the Lambert $W$ function \cite{corless1996}. Rewriting \eqref{eq:engain} in terms of the pulse gives
\begin{align}
	G(A) &= \left( \frac{\Es}{E} W \left( \frac{E}{\Es} \textrm{e}^{E/\Es} \textrm{e}^{g_0 L_g} \right) \right)^{1/2} A
\end{align}
for the amplification of the pulse within the Er-doped fibre. Repeating this process for the loss, dispersion, and nonlinearity yields
\begin{align}
	L(A) &= (1 - R) \textrm{e}^{- \alpha L_T}A, \\
	D(A) &= \FTi{\textrm{e}^{i \omega^2 L_D\beta_2/2} \FT{A}}, \\
	F(A) &= A \textrm{e}^{i \gamma |A|^2 L_f},
\end{align}
where $R$ is the reflectivity of the output coupler (depending on the layout of the laser cavity the loss may instead take the form $L(A) = R \textrm{e}^{- \alpha L_T}A$), $L_T$ is the total length of the laser circuit, $L_D$ is the length of the dispersive medium, $L_f$ is the length of fibre between the amplifier and the output coupler, and where $\mathcal{F}$ denotes the Fourier transform. Finally, we must assume a form for the modulation. The modulation is considered to be applied externally in which ever way the operator sees fit. For simplicity the representation is taken as the Gaussian
\begin{align}
	M(A) &= \textrm{e}^{-T^2 / 2 T_M^2} A,
	\label{eq:modform}
\end{align}
where $T_M$ is the characteristic width of the modulation. This assumption is not particularly restrictive. Calcaterra and Boldt have shown that Gaussians can form a basis for $L^2(\mathbb{R})$ \cite{calcaterra2008a}. Therefore, we can approximate any modulation function in $L^2(\mathbb{R})$ by a sum of \eqref{eq:modform}.

@@

Now that we have the algebraic effect of each section of the cavity, we are ready to compose each map together to give the effect of one round trip of the cavity. Thus, we must now consider the order of the components. As we are most interested in the output of the laser cavity, we shall start with the loss. We then pass the pulse through the CFBG followed by the modulator. The pulse then travels through the Er-doped fibre to be amplified, and finally we consider the effect of the nonlinearity. The nonlinearity must immediately follow the gain---this is the section of the cavity where the pulse is most energetic, and hence, has the largest effect. Note, the permutation of the other components is indeed important as the operators do not, in general, commute---this is in contrast to the models described in Section \ref{sec:discrete}. Moreover, we prefer the loss to follow the gain and nonlinearity as we wish to minimize the length over which the nonlinearity has an effect, and maximize the output power. Functionally we denote one round trip of the cavity by
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A))))).
	\label{eq:order}
\end{align}
 The pulse after one complete circuit of the laser cavity is then returned back into the cavity to restart the process. A steady solution to this model is one in which the envelope and chirp are unchanged after traversing every component in the cavity---we are uninterested in the phase. That is, such that $\mathcal{L}(A) = A \textrm{e}^{i \phi}$---for some $\phi \in \mathbb{R}$. An example of the evolution of the envelope during one round trip of the laser cavity can be found in Figure \ref{fig:cavityevo}.

\section{Results}
\label{sec:results}
We split the results into two subsections. In the following subsection we investigate the low nonlinearity limit, and in Section \ref{sec:nlresults} we look at the full nonlinear model.

@@

\begin{figure}[tbp]
	\centering
	\includegraphics{Sample_Gauss}
	\caption{Solution to the linear model given by \eqref{eq:A0}.}
	\label{fig:samplegauss}
\end{figure}

@@

\subsection{Nonlinear Solution and Instability}
\label{sec:nlresults}
In this iterative scheme---as well as within the laboratory---we must specify the input pulse shape. The most common form is a hyperbolic secant \cite{coen1997, finot2008, mitschke1986, rothenberg1989b, tomlinson1984}, which we assume has the exact form
\begin{align}
	A_0 = \Gamma \sech \left( 2 T \right) \textrm{e}^{i \pi / 4},
	\label{eq:nlA0}
\end{align}
where $\Gamma$ is a normalizing factor chosen so the pulse has an initial energy of $E_0 = 0.1$, unless specified otherwise.

\begin{figure}[tbp]
	\centering
	\includegraphics{ChirpFT}
	\caption{Chirp and Fourier transform in equilibrium state obtained by \eqref{eq:nlA0}, with values $a = 8 \times 10^3$, $h = 0.04$, $b = 1$, $s = 0.2$, and $E_0 = 0.1$.}
	\label{fig:chirpft}
\end{figure}

In Figure \ref{fig:chirpft} we show the chirp and Fourier transform of the pulse \eqref{eq:nlA0} once it has reached equilibrium for the nominal parameter values given in \eqref{eq:ndparam}. In this nonlinear case, we find that the equilibrium pulse is no longer Gaussian---as evidenced by the Fourier transform. Notice that compared to the linear case, higher frequency modes are introduced due to the nonlinearity. Moreover, we find the pulse is linearly chirped near the peak, but, in the tails, $|T| > 1$, the chirp begins to saturate---consistent with experimental results \cite{chen2008, rothenberg1989b, tomlinson1985}. As the nonlinearity parameter, $b$, is increased, the two lobes in the Fourier transform become more pronounced. However, once a critical point is reached SPM begins to play a more substantial factor, leading to modulation instability. This phenomenon is highlighted in Figure \ref{fig:breakevo}. By increasing the nonlinearity, and decreasing the dispersion we find the pulse begins to `breathe'. During the first two dozen round trips of the cavity, the SPM compounds and becomes too parasitic to the pulse. This in turn induces modulation instability, and degrades the pulse, making the pulse no longer stable or sustainable. On the other hand, having a more moderate value of $b$, the pulse is able to equilibrate even with a less favourable initial pulse (Figure \ref{fig:convevo}). The pulse is able to shed the initial left lobe because of dispersion and modulation, and the central lobe is able to grow due to the gain medium. This allows the pulse to come to equilibrium very quickly. Furthermore, since the intensity of the initial pulse is relatively small, the effect of SPM is negligible during the first few round trips which allows the laser to select preferable modes. These preferable modes then get amplified which stabilizes the pulse. From this, it is clear the initial shape of the pulse is much less important than the initial energy of the pulse, $E_0$, or the strength of the nonlinearity, $b$.

\begin{figure}[tbp]
	\centering
	\includegraphics{Break}
	\caption{Example of a pulse destabilizing from the initial pulse \eqref{eq:nlA0} ($a = 8 \times 10^3$, $h = 0.04$, $b = 1.6$, $s = 0.1$, and $E_0 = 0.1$).}
	\label{fig:breakevo}
\end{figure}

\begin{figure}[tbp]
	\centering
	\includegraphics{Conv}
	\caption{Example of a pulse coming to equilibrium from noise ($a = 8 \times 10^3$, $h = 0.04$, $b = 1.0$, $s = 0.1$, and $E_0 = 0.1$).}
	\label{fig:convevo}
\end{figure}

We now turn our attention to characterizing the stability of a laser by exploring the parameter space. We focus on the $s$--$b$ plane in particular since $a$ and $h$ effect only the amplitude of the pulse and do not contribute to the interplay of dispersion and the nonlinearity. Additionally, a conservative initial energy will effect the dynamics much less than a more intense pulse since the nonlinearity will be negligible while coming to equilibrium. To identify the stability of a pulse we examine the relative change in the pulse's envelope between consecutive iterations. We compute this error by
\begin{align}
	\textrm{E} = \frac{\| |A_i| - |A_{i-1}| \|_2}{\| A_{i-1} \|_2},
	\label{eq:error}
\end{align}
where $\| \cdot \|_2$ denotes the $L^2(\mathbb{R})$ norm, and $i$ is chosen to be sufficiently large to guarantee the pulse reaches equilibrium or degrades due to modulation instability---we use $i = 250$ in our experiments (Figure \ref{fig:error}). Note that we take the modulus of the pulse since we are only interested in the evolution of the envelope. We find a very sharp and rich boundary between two regions. In the upper-left region the error is $\bigO{1}$, this is where the nonlinearity is too strong, and the SPM induces modulation instability. In the lower-right region we find the opposite behaviour---the pulse reaches an equilibrium state and is stable. The error is on the order of machine precision, or slightly higher due to approximating a smooth continuous function by discrete points.

\begin{figure}[tbp]
	\centering
	\includegraphics{ParamSpaceErr}
	\caption{Relative error of a pulse's envelope between round trips 249 and 250 ($a = 8 \times 10^3$, $h = 0.04$, and $E_0 = 0.1$).}
	\label{fig:error}
\end{figure}

Also of interest is the energy of the pulse at equilibrium. We compute the energy using \eqref{eq:energy}, or equivalently 
\begin{align}
	E &= \| A_i \|_2,
	\label{eq:energy2norm}
\end{align}
where, again, $i = 250$ in our experiments. Recall from \eqref{eq:order}, this energy is the immediately before passing through the output coupler, and so the energy of each pulse emitted by the laser is instead $(1 - h^2) \| A_i \|_2$, and $h^2 \| A_i \|_2$ remains in the cavity. We plot \eqref{eq:energy2norm} in Figure \ref{fig:energy}. Unsurprisingly, we find the same sharp boundary as we saw in Figure \ref{fig:error}. In the unstable region (upper-left) the energy is relatively small. Since the pulse is not stable in this region it is not able to foster prominent modes, thus, the intensity is low across the entire pulse. In the stable region (lower-right) we find that the energy smoothly decays as both $b$ and $s$ increase, with the contours being hyperbolic-esque. From this it is clear that a more energetic laser requires weaker dispersion, however, the nonlinearity must be correspondingly low, otherwise modulation instability will destroy the pulse.

\begin{figure}[tbp]
	\centering
	\includegraphics{ParamSpaceEnergy}
	\caption{Energy of the pulse after 250 round trips ($a = 8 \times 10^3$, $h = 0.04$, and $E_0 = 0.1$).}
	\label{fig:energy}
\end{figure}

Finally, we briefly look at how the order of the components effects the dynamics. In Section \ref{sec:effects} we justified the order of the gain, nonlinearity, and loss, but, just assumed dispersion preceded modulation. Here we consider the effect of swapping the dispersion and modulation components. The energy landscape (Figure \ref{fig:energyswitch}) is as we would expect. In the previous case of having modulation following dispersion, more energy is removed from the pulse tails. Whereas, having modulation first, more of this energy remains in the cavity. This has the effect of making the laser more powerful, which in turn increases the unstable region. Comparing with Figure \ref{fig:energy}, this is what we find. The unstable upper-left region spreads farther rightward, and the energy is larger in the stable region---especially farther away from the origin.

\begin{figure}[tbp]
	\centering
	\includegraphics{ParamSpaceEnergySwitch}
	\caption{Energy of the pulse after 250 round trips with dispersion and modulation swapped ($a = 8 \times 10^3$, $h = 0.04$, and $E_0 = 0.1$).}
	\label{fig:energyswitch}
\end{figure}

\section{Conclusion}
\label{sec:conclusion}

\section{Acknowledgement}
BM acknowledges the support provided by the EPSRC Centre for Doctoral Training in Industrially Focused Mathematical Modelling (EP/L015803/1).
